Evolutionary rates of SARS-CoV-2

### 2022-08-06 manuscript/variant_rates.tex
split fitness scipt

\begin{figure*}[tb]
    \includegraphics[width=\textwidth]{figures/rtt/20I_rtt.png}
    \caption{{\bf Divergence increases linearly with time (20I (Alpha)).} Each panel shows the number of within-clade mutations (total (A), amino acid changing (B), synonymous (C)) as a function of time.
    The blue line in panel A indicates the divergence cut-off, panels B\&C only show sequences that pass the divergence filter. Each panel also shows mean $\pm$ standard deviation and a weighted linear fit. Analogous figures for all clades considered are included in the appendix.
    \label{fig:rate_alpha}}
\end{figure*}

@@

Above, we found that the rate of synonymous mutations is comparable within and between clades without any striking signal of temporal variation.
This is expected, as synonymous positions are rarely a locus of adaptation and tend to have a small effects on fitness in large parts of the genomes of RNA viruses \citep{zanini_vivo_2017} (outside of specific regions with important RNA elements).
To assess the level of constraint on the SARS-CoV-2 genome, we made use of the ``rare mutations'' annotation by Nextclade.
Nextclade attaches each sequence to a reference tree and determines by which mutations it differs from the attachment point.
For each pango lineage (as determined by Nextclade), we could how often these ``rare mutations'' are observed, normalize this against the relative rate of mutation of the reference nucleotide (see Supp.~Fig.~\ref{fig:mutation_distribution}), and use this as a proxy for mutational tolerance.

>>>>

\begin{figure*}[tb]
    \includegraphics[width=\textwidth]{figures/rtt/20I_rtt.pdf}
    \caption{{\bf Divergence increases linearly with time (20I (Alpha)).} Each panel shows the number of within-clade mutations (total (A), amino acid changing (B), synonymous (C)) as a function of time.
    The blue line in panel A indicates the divergence cut-off, panels B\&C only show sequences that pass the divergence filter. Each panel also shows mean $\pm$ standard deviation and a weighted linear fit. Analogous figures for all clades considered are included in the appendix.
    \label{fig:rate_alpha}}
\end{figure*}

@@

Above, we found that the rate of synonymous mutations is comparable within and between clades without any striking signal of temporal variation.
This is expected, as synonymous positions are rarely a locus of adaptation and tend to have a small effects on fitness in large parts of the genomes of RNA viruses \citep{zanini_vivo_2017} (outside of specific regions with important RNA elements).
To assess the level of constraint on the SARS-CoV-2 genome, we made use of the ``rare mutations'' annotation by Nextclade.
Nextclade attaches each sequence to a reference tree and determines by which mutations it differs from the attachment point.
For each pango lineage (as determined by Nextclade), we count how often these ``rare mutations'' are observed, normalize this against the relative rate of mutation of the reference nucleotide (see Supp.~Fig.~\ref{fig:mutation_distribution}), and use this as a proxy for mutational tolerance.
